---
title: "Data joining"
author: "Naomi"
date: "2025-04-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This imports the cleaned dataset from the initial data creation step in Data Creation.Rmd
```{r}
health <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQeaKrReyifqjPVoWB2pYHAo54wLlybjEnMZiPP0i6tdpWb-RLqmwrCvhz8HVOBLCjEo_0qYmrnOkEo/pub?output=csv")

```


We now want to add the population level to the dataset
```{r}
library(readr)
library(dplyr)
library(tidycensus)
library(tidyverse)

# Get 2019 ACS 5-Year population estimates for all US places
city_pop <- get_acs(
  geography = "place",
  variables = "B01003_001",  # Total population
  year = 2019,
  survey = "acs5",
  output = "wide",
  geometry = FALSE
)

# This converts the columns to a more readable format
top_cities <- city_pop %>%
  select(GEOID, NAME, B01003_001E) %>%  # GEOID is the FIPS code
  rename(
    fips = GEOID,
    city = NAME,
    population = B01003_001E
  ) %>%
  arrange(desc(population))

# This ensures when we join our data the columns in health and top_cities have the same variable name
top_cities <- rename(top_cities, CityFIPS = fips)

#This casts top_cities as numeric so that it can be joined with the numeric CityFIPS column in health
top_cities$CityFIPS <- as.numeric(top_cities$CityFIPS)

# This is a slightly destructive join
# It removes two cities from our original dataset because their FIPS are not in the ACS dataset
# The fact that they are not in the ACS dataset means we probably don't care about them
# And 498 cities is an acceptable number of datapoints for the scope of this project
healthwfips <- inner_join(health, top_cities, by = "CityFIPS")

```

The questions we want to consider are as such:

Consider the variables:
- mental illness
- binge drinking
- asthma
- arthritis
- cancer
- kidney disease
- high blood pressure
- diabetes
- COPD


Group By State
```{r}
# The population represented here is the sum of the considered cities
# It represents a SAMPLE SIZE and is NOT IS NOT representative of the total population of those states
state_means <- healthwfips %>% group_by(StateDesc) %>%
  summarise(avgMentalIllness = mean(MHLTH), avgBingeDrinking = mean(BINGE), avgChildhoodAsthma = mean(CASTHMA), avgArthritis = mean(ARTHRITIS), avgCancer = mean(CANCER), avgKidneyDisease = mean(KIDNEY), avgHighBP = mean(BPHIGH), avgDiabetes = mean(DIABETES), avgCOPD = mean(COPD), statePopulation = sum(population))

state_means <- state_means %>%
  mutate(StateDesc = tolower(StateDesc))

state_means$StateDesc[state_means$StateDesc == "south carolin"] <- "south carolina"
state_means$StateDesc[state_means$StateDesc == "north carolin"] <- "north carolina"

```



- arthritis
- kidney disease
- high blood pressure
- diabetes

Map By State
```{r}
install.packages("viridis")
library(ggplot2)
library(dplyr)
library(maps)      # for map_data
library(viridis)   # optional: better color scale


my_data <- state_means %>% mutate(state = tolower(StateDesc))
map_data_combined <- left_join(states_map, my_data, by = c("region" = "state"))

ggplot(map_data_combined, aes(x = long, y = lat, group = group, fill = avgMentalIllness)) +
  geom_polygon(color = "white", size = 0.3) +
  coord_fixed(1.3) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Mental Illness Rate"
  ) +
  labs(
    title = "Average Mental Illness Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )

ggplot(map_data_combined, aes(x = long, y = lat, group = group, fill = avgBingeDrinking)) +
  geom_polygon(color = "white", size = 0.3) +
  coord_fixed(1.3) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Binge Drinking Rate"
  ) +
  labs(
    title = "Average Binge Drinking Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )

ggplot(map_data_combined, aes(x = long, y = lat, group = group, fill = avgChildhoodAsthma)) +
  geom_polygon(color = "white", size = 0.3) +
  coord_fixed(1.3) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Childhood Asthma Rate"
  ) +
  labs(
    title = "Average Childhood Asthma Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )

ggplot(map_data_combined, aes(x = long, y = lat, group = group, fill = avgArthritis)) +
  geom_polygon(color = "white", size = 0.3) +
  coord_fixed(1.3) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Arthritis Rate"
  ) +
  labs(
    title = "Average Arthritis Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )

ggplot(map_data_combined, aes(x = long, y = lat, group = group, fill = avgKidneyDisease)) +
  geom_polygon(color = "white", size = 0.3) +
  coord_fixed(1.3) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Kidney Disease Rate"
  ) +
  labs(
    title = "Average Kidney Disease Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )


ggplot(map_data_combined, aes(x = long, y = lat, group = group, fill = avgHighBP)) +
  geom_polygon(color = "white", size = 0.3) +
  coord_fixed(1.3) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg High Blood Pressure Rate"
  ) +
  labs(
    title = "Average High Blood Pressure Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )


ggplot(map_data_combined, aes(x = long, y = lat, group = group, fill = avgCancer)) +
  geom_polygon(color = "white", size = 0.3) +
  coord_fixed(1.3) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Cancer Rate"
  ) +
  labs(
    title = "Average Cancer Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8))

ggplot(map_data_combined, aes(x = long, y = lat, group = group, fill = avgDiabetes)) +
  geom_polygon(color = "white", size = 0.3) +
  coord_fixed(1.3) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Diabetes Rate"
  ) +
  labs(
    title = "Average Diabetes Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8))


ggplot(map_data_combined, aes(x = long, y = lat, group = group, fill = avgCOPD)) +
  geom_polygon(color = "white", size = 0.3) +
  coord_fixed(1.3) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg COPD Rate"
  ) +
  labs(
    title = "Average COPD Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )



```




