---
title: "DS2020 Final Project"
output: github document
date: "`r Sys.Date()`"
author: "Grace Wu, Naomi Mauss"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## DS2020 Final Project: Health Data in 500 Cities

Grace Wu, Naomi Mauss

### Data

#### Description of Dataset:

-   This dataset includes information regarding chronic diseases related to three categories (unhalthy behaviors, health outcomes, and prevention). The data was provided by the CDC and consists of 500 of the largest cities.

```{r}
Health <- read.csv("C:\\Users\\songb\\Downloads\\500_Cities__Local_Data_for_Better_Health__2019_release_20250331.csv")
```

#### Data Cleaning
```{r}

Health <- Health[complete.cases(Health), ]

```

#### Marginal Summaries
```{r}
head(Health)
dim(Health)
str(Health)
summary(Health)
```

#### Questions to be addressed

Consider the variables:
- mental illness
- binge drinking
- asthma
- arthritis
- cancer
- kidney disease
- high blood pressure
- diabetes
- COPD

When grouped by state, what states have the highest rates of each variable?
When grouped by state, what states have the lowest rates of each variable?

Which places have the highest rate of each variable?
Which places have the lowest rate of each variable?

Are there any trends regarding the least healthy places to live in the United States?
Are there any trends regarding the healthiest places to live in the United States?

The goal of this project is to determine where in the United States is the healthiest, and where is the least healthy. This information is important for people who might want to know how the places they move might affect their health.

