---
title: "README"
author: "Naomi"
date: "2025-04-16"
output: github_document
format:
  html:
    df-print: gt
    page-layout: full
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This imports the cleaned dataset from the initial data creation step in
Data Creation.Rmd

```{r}
health <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQeaKrReyifqjPVoWB2pYHAo54wLlybjEnMZiPP0i6tdpWb-RLqmwrCvhz8HVOBLCjEo_0qYmrnOkEo/pub?output=csv")


healthwfips <-read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTZ4PRD5X0Zcr1rEWwCTusXmef1IkAFgnpPXeJ1dZ3lVpp7iX-ZsbMFCOpkzS-zSA/pub?output=csv")


```

```{R}

library(readr)
library(dplyr)
library(tidycensus)
library(tidyverse)
library(data.table)
library(gt)
library(xgboost)
library(tigris)
library(ggplot2)
library(sf)
library(stringr)
library(rmapshaper)
library(knitr)

```





``` {R}

state_means <- healthwfips %>%
  group_by(StateDesc) %>%
  summarise(
    avgMentalIllness = mean(MHLTH, na.rm = TRUE),
    avgBingeDrinking = mean(BINGE, na.rm = TRUE),
    avgChildhoodAsthma = mean(CASTHMA, na.rm = TRUE),
    avgArthritis = mean(ARTHRITIS, na.rm = TRUE),
    avgCancer = mean(CANCER, na.rm = TRUE),
    avgKidneyDisease = mean(KIDNEY, na.rm = TRUE),
    avgHighBP = mean(BPHIGH, na.rm = TRUE),
    avgDiabetes = mean(DIABETES, na.rm = TRUE),
    avgCOPD = mean(COPD, na.rm = TRUE),
    .groups = "drop"
  )

state_means <- state_means %>%
  mutate(StateDesc = tolower(StateDesc))

state_means <- healthwfips %>% group_by(StateDesc) %>%
  summarise(avgMentalIllness = mean(MHLTH), avgBingeDrinking = mean(BINGE), avgChildhoodAsthma = mean(CASTHMA), avgArthritis = mean(ARTHRITIS), avgCancer = mean(CANCER), avgKidneyDisease = mean(KIDNEY), avgHighBP = mean(BPHIGH), avgDiabetes = mean(DIABETES), avgCOPD = mean(COPD))




state_means$StateDesc[state_means$StateDesc == "South Carolin"] <- "South Carolina"
state_means$StateDesc[state_means$StateDesc == "North Carolin"] <- "North Carolina"

```


Prepare the data to map by state

```{r}
#install.packages("viridis")
#library(ggplot2)
#library(dplyr)
#library(maps)      # for map_data
#library(viridis)   # optional: better color scale

# state_means <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRPtd9hIIamvljlqlmKcpm8xzcoSxkiNT4bLA78cgczFQqrSjI1DRrKnNiKR5Ger9CZHL8AEd0bx7OW/pub?output=csv")

state_means$StateDesc <- str_to_title(state_means$StateDesc)

```

```{R}
# Download the shapefile for all states, including Alaska and Hawaii
us_states_sf <- states(cb = TRUE)

# View the structure of the data (optional)
head(us_states_sf)

# Map state names in your data to their respective state abbreviations
state_abbreviations <- c(
  "Alaska" = "AK", "Hawaii" = "HI", "Alabama" = "AL", "Arizona" = "AZ", 
  "Arkansas" = "AR", "California" = "CA", "Colorado" = "CO", "Connecticut" = "CT",
  "Delaware" = "DE", "Florida" = "FL", "Georgia" = "GA", "Hawaii" = "HI", 
  "Idaho" = "ID", "Illinois" = "IL", "Indiana" = "IN", "Iowa" = "IA", 
  "Kansas" = "KS", "Kentucky" = "KY", "Louisiana" = "LA", "Maine" = "ME", 
  "Maryland" = "MD", "Massachusetts" = "MA", "Michigan" = "MI", "Minnesota" = "MN", 
  "Mississippi" = "MS", "Missouri" = "MO", "Montana" = "MT", "Nebraska" = "NE", 
  "Nevada" = "NV", "New Hampshire" = "NH", "New Jersey" = "NJ", "New Mexico" = "NM", 
  "New York" = "NY", "North Carolina" = "NC", "North Dakota" = "ND", "Ohio" = "OH", 
  "Oklahoma" = "OK", "Oregon" = "OR", "Pennsylvania" = "PA", "Rhode Island" = "RI", 
  "South Carolina" = "SC", "South Dakota" = "SD", "Tennessee" = "TN", "Texas" = "TX", 
  "Utah" = "UT", "Vermont" = "VT", "Virginia" = "VA", "Washington" = "WA", 
  "West Virginia" = "WV", "Wisconsin" = "WI", "Wyoming" = "WY"
)

# my_data <- state_means %>%
#   mutate(state = tolower(StateDesc)) %>%
#   mutate(state_abbr = recode(state, !!!state_abbreviations))

# Merge map data with your data
map_data_combined <- left_join(us_states_sf, state_means, by = c("NAME" = "StateDesc"))

```

```{R}
# install.packages("writexl")
 
# library(writexl)
 
# write_xlsx(map_data_combined, 'C:\\Users\\songb\\Desktop\\ndata.xlsx')
 
 
# write_xlsx(state_means, 'C:\\Users\\songb\\Desktop\\state_means.xlsx')



# state_means[state_means == 'district of c'] <- 'DC'

```



Now we make the maps and box plots


Mental Illness Map
```{R}

# Plot full map with AK/HI shifted
formatting <- tigris::states() %>%
  dplyr::filter(GEOID < "60") %>%
  tigris::shift_geometry()

formatting <- formatting %>%
  left_join(state_means, by = c("NAME" = "StateDesc"))

# plot(formatting$geometry, formatting$avgArthritis)


ggplot(formatting) +
  geom_sf(aes(fill = avgMentalIllness)) +
  scale_fill_viridis_c(
      option = "plasma",
      direction = -1,
      name = "Avg Mental Illness Rate"
    ) +
    labs(
      title = "Average Mental Illness Rate by State",
      subtitle = "Data from CDC PLACES and ACS",
      caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
    ) +
    theme_void() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12),
      plot.caption = element_text(hjust = 1, size = 8),
      legend.position = "right",
      legend.title = element_text(size = 10, face = "bold"),
      legend.text = element_text(size = 8)
   )


```


``` {html}
# # Download the shapefile for all states, including Alaska and Hawaii
# us_states <- states(cb = TRUE)
# 
# # Convert to sf object
# us_states_sf <- st_as_sf(us_states)
# 
# # Map state names in your data to their respective state abbreviations
# state_abbreviations <- c(
#   "alaska" = "AK", "hawaii" = "HI", "alabama" = "AL", "arizona" = "AZ", 
#   "arkansas" = "AR", "california" = "CA", "colorado" = "CO", "connecticut" = "CT",
#   "delaware" = "DE", "florida" = "FL", "georgia" = "GA", "idaho" = "ID", 
#   "illinois" = "IL", "indiana" = "IN", "iowa" = "IA", "kansas" = "KS", 
#   "kentucky" = "KY", "louisiana" = "LA", "maine" = "ME", "maryland" = "MD", 
#   "massachusetts" = "MA", "michigan" = "MI", "minnesota" = "MN", "mississippi" = "MS", 
#   "missouri" = "MO", "montana" = "MT", "nebraska" = "NE", "nevada" = "NV", 
#   "new hampshire" = "NH", "new jersey" = "NJ", "new mexico" = "NM", 
#   "new york" = "NY", "north carolina" = "NC", "north dakota" = "ND", "ohio" = "OH", 
#   "oklahoma" = "OK", "oregon" = "OR", "pennsylvania" = "PA", "rhode island" = "RI", 
#   "south carolina" = "SC", "south dakota" = "SD", "tennessee" = "TN", "texas" = "TX", 
#   "utah" = "UT", "vermont" = "VT", "virginia" = "VA", "washington" = "WA", 
#   "west virginia" = "WV", "wisconsin" = "WI", "wyoming" = "WY"
# )
# 
# # Merge the map data with your data for the mainland U.S.
# my_data <- state_means %>%
#   mutate(state = tolower(StateDesc)) %>%
#   mutate(state_abbr = recode(state, !!!state_abbreviations))
# 
# map_data_mainland <- us_states_sf %>%
#   filter(!STUSPS %in% c("AK", "HI")) %>%
#   left_join(my_data, by = c(StateDesc = "state_abbr"))
# 
# us_states_shifted <- shift_geometry(
#   us_states_sf,
#   geoid_column = NULL,
#   preserve_area = FALSE,
#   position = c("below", "outside")
# )
# 
# map_data_shifted <- us_states_shifted %>%
#   left_join(state_means, by = c("STUSPS" = "state_abbr"))
# 
# # Plot the map
# ggplot() + 
#   # Mainland U.S. plot
#   geom_sf(data = map_data_mainland, aes(fill = avgMentalIllness)) + 
#   coord_sf(
#     xlim = c(-125, -66),  # Longitude range for mainland U.S.
#     ylim = c(24, 50),      # Latitude range for mainland U.S.
#     expand = FALSE
#   ) + 
#   scale_fill_viridis_c(
#     option = "plasma",
#     direction = -1,
#     name = "Avg Mental Illness Rate"
#   ) +
#   labs(
#     title = "Average Mental Illness Rate by State",
#     subtitle = "Data from CDC PLACES and ACS",
#     caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
#   ) +
#   theme_void() +
#   theme(
#     plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
#     plot.subtitle = element_text(hjust = 0.5, size = 12),
#     plot.caption = element_text(hjust = 1, size = 8),
#     legend.position = "right",
#     legend.title = element_text(size = 10, face = "bold"),
#     legend.text = element_text(size = 8)
#   )







# ggplot(map_data_combined, aes(x = long, y = lat, group = group, fill = avgMentalIllness)) +
#   geom_polygon(color = "white", size = 0.3) +
#   coord_fixed(1.3) +
#   scale_fill_viridis_c(
#     option = "plasma",
#     direction = -1,
#     name = "Avg Mental Illness Rate"
#   ) +
#   labs(
#     title = "Average Mental Illness Rate by State",
#     subtitle = "Data from CDC PLACES and ACS",
#     caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
#   ) +
#   theme_void() +
#   theme(
#     plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
#     plot.subtitle = element_text(hjust = 0.5, size = 12),
#     plot.caption = element_text(hjust = 1, size = 8),
#     legend.position = "right",
#     legend.title = element_text(size = 10, face = "bold"),
#     legend.text = element_text(size = 8))
#     
# ggplot(map_data_combined) + 
#   geom_sf(aes(fill = avgMentalIllness)) +  # Fill by 'value' column from your data
#   coord_sf(
#     xlim = c(-125, -66),  # Longitude range to zoom in on the mainland US
#     ylim = c(24, 50),      # Latitude range for mainland U.S.
#     expand = FALSE
#   ) + 
#   scale_fill_viridis_c(
#     option = "plasma",
#     direction = -1,
#     name = "Avg Mental Illness Rate"
#   ) +
#   labs(
#     title = "Average Mental Illness Rate by State",
#     subtitle = "Data from CDC PLACES and ACS",
#     caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
#   ) +
#   theme_void() +
#   theme(
#     plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
#     plot.subtitle = element_text(hjust = 0.5, size = 12),
#     plot.caption = element_text(hjust = 1, size = 8),
#     legend.position = "right",
#     legend.title = element_text(size = 10, face = "bold"),
#     legend.text = element_text(size = 8)
#   )
    
```


Mental Illness Box Plot Desc

```{R}
stored_means <- data.frame(health %>%
  group_by(StateDesc) %>%
  summarise(mean_COPD = mean(MHLTH, na.rm = TRUE)) %>%
  arrange(mean_COPD))

highlight_states <- c(
  head(stored_means$StateDesc, 5),
  tail(stored_means$StateDesc, 5)
)

health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"))

health$StateDesc <- reorder(health$StateDesc, health$MHLTH, FUN = median)


health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"),
         StateDesc = factor(StateDesc, levels = stored_means$StateDesc))  # Ensure the correct order

x_mapping <- health %>%
  filter(Highlight == "Highlight") %>%
  arrange(COPD) %>%
  mutate(x_pos = as.numeric(factor(StateDesc)))

# Plot
ggplot() +
  # Plot all "Other" states as points (mean) in the middle
  stat_summary(data = filter(health, Highlight == "Other"),
               aes(x = StateDesc, y = MHLTH),
               fun = mean, geom = "point", shape = 21, fill = "gray60", size = 2) +

  # Plot boxplots for highlighted states (top and bottom 5)
  geom_boxplot(data = filter(health, Highlight == "Highlight"),
               aes(x = StateDesc, y = MHLTH),
               width = 0.8, fill = "steelblue", color = "black") +

  scale_x_continuous(
    breaks = x_mapping$x_pos[x_mapping$StateDesc %in% highlight_states],
    labels = x_mapping$StateDesc[x_mapping$StateDesc %in% highlight_states]) +
  
  labs(title = "Mental Illness Rates by City: Highlighting Extremes by State",
       x = "State (ordered by median mental illness)",
       y = "Mental Illness Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1)) +
  scale_x_discrete(limits = levels(health$StateDesc), labels = function(x) ifelse(x %in% highlight_states, x, ""))

```



Binge Drinking Map
```{R}
ggplot(formatting) +
  geom_sf(aes(fill = avgBingeDrinking))+
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Binge Drinking Rate"
  ) +
  labs(
    title = "Average Binge Drinking Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )
```


Binge Drinking Box Plot Desc

```{R}
stored_means <- health %>%
  group_by(StateDesc) %>%
  summarise(mean_COPD = mean(BINGE, na.rm = TRUE)) %>%
  arrange(mean_COPD)

highlight_states <- c(
  head(stored_means$StateDesc, 5),
  tail(stored_means$StateDesc, 5)
)

health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"))

health$StateDesc <- reorder(health$StateDesc, health$BINGE, FUN = median)


health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"),
         StateDesc = factor(StateDesc, levels = stored_means$StateDesc))  # Ensure the correct order

x_mapping <- health %>%
  filter(Highlight == "Highlight") %>%
  arrange(COPD) %>%
  mutate(x_pos = as.numeric(factor(StateDesc)))

# Plot
ggplot() +
  # Plot all "Other" states as points (mean) in the middle
  stat_summary(data = filter(health, Highlight == "Other"),
               aes(x = StateDesc, y = BINGE),
               fun = mean, geom = "point", shape = 21, fill = "gray60", size = 2) +

  # Plot boxplots for highlighted states (top and bottom 5)
  geom_boxplot(data = filter(health, Highlight == "Highlight"),
               aes(x = StateDesc, y = BINGE),
               width = 0.8, fill = "steelblue", color = "black") +

  scale_x_continuous(
    breaks = x_mapping$x_pos[x_mapping$StateDesc %in% highlight_states],
    labels = x_mapping$StateDesc[x_mapping$StateDesc %in% highlight_states]) +
  
  labs(title = "Binge Drinking Rates by City: Highlighting Extremes by State",
       x = "State (ordered by median binge drinking)",
       y = "Binge Drinking Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1)) +
  scale_x_discrete(limits = levels(health$StateDesc), labels = function(x) ifelse(x %in% highlight_states, x, ""))

```




Asthma Map

```{R}
ggplot(formatting) +
  geom_sf(aes(fill = avgChildhoodAsthma)) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Childhood Asthma Rate"
  ) +
  labs(
    title = "Average Childhood Asthma Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )
```


Childhood Asthma Box Plot Desc

```{R}
stored_means <- health %>%
  group_by(StateDesc) %>%
  summarise(mean_COPD = mean(CASTHMA, na.rm = TRUE)) %>%
  arrange(mean_COPD)

highlight_states <- c(
  head(stored_means$StateDesc, 5),
  tail(stored_means$StateDesc, 5)
)

health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"))

health$StateDesc <- reorder(health$StateDesc, health$CASTHMA, FUN = median)


health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"),
         StateDesc = factor(StateDesc, levels = stored_means$StateDesc))  # Ensure the correct order

x_mapping <- health %>%
  filter(Highlight == "Highlight") %>%
  arrange(COPD) %>%
  mutate(x_pos = as.numeric(factor(StateDesc)))

# Plot
ggplot() +
  # Plot all "Other" states as points (mean) in the middle
  stat_summary(data = filter(health, Highlight == "Other"),
               aes(x = StateDesc, y = CASTHMA),
               fun = mean, geom = "point", shape = 21, fill = "gray60", size = 2) +

  # Plot boxplots for highlighted states (top and bottom 5)
  geom_boxplot(data = filter(health, Highlight == "Highlight"),
               aes(x = StateDesc, y = CASTHMA),
               width = 0.8, fill = "steelblue", color = "black") +

  scale_x_continuous(
    breaks = x_mapping$x_pos[x_mapping$StateDesc %in% highlight_states],
    labels = x_mapping$StateDesc[x_mapping$StateDesc %in% highlight_states]) +
  
  labs(title = "Childhood Asthma Rates by City: Highlighting Extremes by State",
       x = "State (ordered by median childhood asthma)",
       y = "Childhood Asthma Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1)) +
  scale_x_discrete(limits = levels(health$StateDesc), labels = function(x) ifelse(x %in% highlight_states, x, ""))

```



Arthritis Map
```{R}
ggplot(formatting) +
  geom_sf(aes(fill = avgArthritis)) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Arthritis Rate"
  ) +
  labs(
    title = "Average Arthritis Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )
```


Arthritis Box Plot Desc

```{R}
stored_means <- health %>%
  group_by(StateDesc) %>%
  summarise(mean_COPD = mean(ARTHRITIS, na.rm = TRUE)) %>%
  arrange(mean_COPD)

highlight_states <- c(
  head(stored_means$StateDesc, 5),
  tail(stored_means$StateDesc, 5)
)

health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"))

health$StateDesc <- reorder(health$StateDesc, health$ARTHRITIS, FUN = median)


health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"),
         StateDesc = factor(StateDesc, levels = stored_means$StateDesc))  # Ensure the correct order

x_mapping <- health %>%
  filter(Highlight == "Highlight") %>%
  arrange(COPD) %>%
  mutate(x_pos = as.numeric(factor(StateDesc)))

# Plot
ggplot() +
  # Plot all "Other" states as points (mean) in the middle
  stat_summary(data = filter(health, Highlight == "Other"),
               aes(x = StateDesc, y = ARTHRITIS),
               fun = mean, geom = "point", shape = 21, fill = "gray60", size = 2) +

  # Plot boxplots for highlighted states (top and bottom 5)
  geom_boxplot(data = filter(health, Highlight == "Highlight"),
               aes(x = StateDesc, y = ARTHRITIS),
               width = 0.8, fill = "steelblue", color = "black") +

  scale_x_continuous(
    breaks = x_mapping$x_pos[x_mapping$StateDesc %in% highlight_states],
    labels = x_mapping$StateDesc[x_mapping$StateDesc %in% highlight_states]) +
  
  labs(title = "Arthritis Rates by City: Highlighting Extremes by State",
       x = "State (ordered by median arthritis)",
       y = "Childhood Arthritis Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1)) +
  scale_x_discrete(limits = levels(health$StateDesc), labels = function(x) ifelse(x %in% highlight_states, x, ""))

```



Kidney Disease Map
```{R}
ggplot(formatting) +
  geom_sf(aes(fill = avgKidneyDisease)) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Kidney Disease Rate"
  ) +
  labs(
    title = "Average Kidney Disease Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )
```


Kidney Disease Box Plot Desc

```{R}
stored_means <- health %>%
  group_by(StateDesc) %>%
  summarise(mean_COPD = mean(KIDNEY, na.rm = TRUE)) %>%
  arrange(mean_COPD)

highlight_states <- c(
  head(stored_means$StateDesc, 5),
  tail(stored_means$StateDesc, 5)
)

health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"))

health$StateDesc <- reorder(health$StateDesc, health$KIDNEY, FUN = median)


health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"),
         StateDesc = factor(StateDesc, levels = stored_means$StateDesc))  # Ensure the correct order

x_mapping <- health %>%
  filter(Highlight == "Highlight") %>%
  arrange(COPD) %>%
  mutate(x_pos = as.numeric(factor(StateDesc)))

# Plot
ggplot() +
  # Plot all "Other" states as points (mean) in the middle
  stat_summary(data = filter(health, Highlight == "Other"),
               aes(x = StateDesc, y = KIDNEY),
               fun = mean, geom = "point", shape = 21, fill = "gray60", size = 2) +

  # Plot boxplots for highlighted states (top and bottom 5)
  geom_boxplot(data = filter(health, Highlight == "Highlight"),
               aes(x = StateDesc, y = KIDNEY),
               width = 0.8, fill = "steelblue", color = "black") +

  scale_x_continuous(
    breaks = x_mapping$x_pos[x_mapping$StateDesc %in% highlight_states],
    labels = x_mapping$StateDesc[x_mapping$StateDesc %in% highlight_states]) +
  
  labs(title = "Kidney Disease Rates by City: Highlighting Extremes by State",
       x = "State (ordered by median kidney disease)",
       y = "Kidney Disease Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1)) +
  scale_x_discrete(limits = levels(health$StateDesc), labels = function(x) ifelse(x %in% highlight_states, x, ""))

```


High Blood Pressure Map
```{R}
ggplot(formatting) +
  geom_sf(aes(fill = avgHighBP)) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg High Blood Pressure Rate"
  ) +
  labs(
    title = "Average High Blood Pressure Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )
```


High Blood Pressure Box Plot Desc

```{R}
stored_means <- health %>%
  group_by(StateDesc) %>%
  summarise(mean_COPD = mean(BPHIGH, na.rm = TRUE)) %>%
  arrange(mean_COPD)

highlight_states <- c(
  head(stored_means$StateDesc, 5),
  tail(stored_means$StateDesc, 5)
)

health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"))

health$StateDesc <- reorder(health$StateDesc, health$BPHIGH, FUN = median)


health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"),
         StateDesc = factor(StateDesc, levels = stored_means$StateDesc))  # Ensure the correct order

x_mapping <- health %>%
  filter(Highlight == "Highlight") %>%
  arrange(COPD) %>%
  mutate(x_pos = as.numeric(factor(StateDesc)))

# Plot
ggplot() +
  # Plot all "Other" states as points (mean) in the middle
  stat_summary(data = filter(health, Highlight == "Other"),
               aes(x = StateDesc, y = BPHIGH),
               fun = mean, geom = "point", shape = 21, fill = "gray60", size = 2) +

  # Plot boxplots for highlighted states (top and bottom 5)
  geom_boxplot(data = filter(health, Highlight == "Highlight"),
               aes(x = StateDesc, y = BPHIGH),
               width = 0.8, fill = "steelblue", color = "black") +

  scale_x_continuous(
    breaks = x_mapping$x_pos[x_mapping$StateDesc %in% highlight_states],
    labels = x_mapping$StateDesc[x_mapping$StateDesc %in% highlight_states]) +
  
  labs(title = "High Blood Pressure Rates by City: Highlighting Extremes by State",
       x = "State (ordered by median high blood pressure)",
       y = "High Blood Pressure Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1)) +
  scale_x_discrete(limits = levels(health$StateDesc), labels = function(x) ifelse(x %in% highlight_states, x, ""))

```


Cancer Map
```{R}
ggplot(formatting) +
  geom_sf(aes(fill = avgCancer)) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Cancer Rate"
  ) +
  labs(
    title = "Average Cancer Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8))

```


Childhood Box Plot Desc

```{R}
stored_means <- health %>%
  group_by(StateDesc) %>%
  summarise(mean_COPD = mean(CANCER, na.rm = TRUE)) %>%
  arrange(mean_COPD)

highlight_states <- c(
  head(stored_means$StateDesc, 5),
  tail(stored_means$StateDesc, 5)
)

health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"))

health$StateDesc <- reorder(health$StateDesc, health$CANCER, FUN = median)


health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"),
         StateDesc = factor(StateDesc, levels = stored_means$StateDesc))  # Ensure the correct order

x_mapping <- health %>%
  filter(Highlight == "Highlight") %>%
  arrange(COPD) %>%
  mutate(x_pos = as.numeric(factor(StateDesc)))

# Plot
ggplot() +
  # Plot all "Other" states as points (mean) in the middle
  stat_summary(data = filter(health, Highlight == "Other"),
               aes(x = StateDesc, y = CANCER),
               fun = mean, geom = "point", shape = 21, fill = "gray60", size = 2) +

  # Plot boxplots for highlighted states (top and bottom 5)
  geom_boxplot(data = filter(health, Highlight == "Highlight"),
               aes(x = StateDesc, y = CANCER),
               width = 0.8, fill = "steelblue", color = "black") +

  scale_x_continuous(
    breaks = x_mapping$x_pos[x_mapping$StateDesc %in% highlight_states],
    labels = x_mapping$StateDesc[x_mapping$StateDesc %in% highlight_states]) +
  
  labs(title = "Cancer Rates by City: Highlighting Extremes by State",
       x = "State (ordered by median cancer)",
       y = "Cancer Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1)) +
  scale_x_discrete(limits = levels(health$StateDesc), labels = function(x) ifelse(x %in% highlight_states, x, ""))

```


Diabetes Map
```{R}

ggplot(formatting) +
  geom_sf(aes(fill = avgDiabetes)) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Diabetes Rate"
  ) +
  labs(
    title = "Average Diabetes Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8))

```


Childhood Box Plot Desc

```{R}
stored_means <- health %>%
  group_by(StateDesc) %>%
  summarise(mean_COPD = mean(DIABETES, na.rm = TRUE)) %>%
  arrange(mean_COPD)

highlight_states <- c(
  head(stored_means$StateDesc, 5),
  tail(stored_means$StateDesc, 5)
)

health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"))

health$StateDesc <- reorder(health$StateDesc, health$DIABETES, FUN = median)


health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"),
         StateDesc = factor(StateDesc, levels = stored_means$StateDesc))  # Ensure the correct order

x_mapping <- health %>%
  filter(Highlight == "Highlight") %>%
  arrange(COPD) %>%
  mutate(x_pos = as.numeric(factor(StateDesc)))

# Plot
ggplot() +
  # Plot all "Other" states as points (mean) in the middle
  stat_summary(data = filter(health, Highlight == "Other"),
               aes(x = StateDesc, y = DIABETES),
               fun = mean, geom = "point", shape = 21, fill = "gray60", size = 2) +

  # Plot boxplots for highlighted states (top and bottom 5)
  geom_boxplot(data = filter(health, Highlight == "Highlight"),
               aes(x = StateDesc, y = DIABETES),
               width = 0.8, fill = "steelblue", color = "black") +

  scale_x_continuous(
    breaks = x_mapping$x_pos[x_mapping$StateDesc %in% highlight_states],
    labels = x_mapping$StateDesc[x_mapping$StateDesc %in% highlight_states]) +
  
  labs(title = "Diabetes Rates by City: Highlighting Extremes by State",
       x = "State (ordered by median diabetes)",
       y = "Diabetes Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1)) +
  scale_x_discrete(limits = levels(health$StateDesc), labels = function(x) ifelse(x %in% highlight_states, x, ""))

```


COPD Map
```{R}
ggplot(formatting) +
  geom_sf(aes(fill = avgCOPD)) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg COPD Rate"
  ) +
  labs(
    title = "Average COPD Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )

```

COPD Box Plot Desc

```{R}
stored_means <- health %>%
  group_by(StateDesc) %>%
  summarise(mean_COPD = mean(COPD, na.rm = TRUE)) %>%
  arrange(mean_COPD)

highlight_states <- c(
  head(stored_means$StateDesc, 5),
  tail(stored_means$StateDesc, 5)
)



health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"))

health$StateDesc <- reorder(health$StateDesc, health$COPD, FUN = median)


health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"),
         StateDesc = factor(StateDesc, levels = stored_means$StateDesc))  # Ensure the correct order

x_mapping <- health %>%
  filter(Highlight == "Highlight") %>%
  arrange(COPD) %>%
  mutate(x_pos = as.numeric(factor(StateDesc)))

# Plot
ggplot() +
  # Plot all "Other" states as points (mean) in the middle
  stat_summary(data = filter(health, Highlight == "Other"),
               aes(x = StateDesc, y = COPD),
               fun = mean, geom = "point", shape = 21, fill = "gray60", size = 2) +

  # Plot boxplots for highlighted states (top and bottom 5)
  geom_boxplot(data = filter(health, Highlight == "Highlight"),
               aes(x = StateDesc, y = COPD),
               width = 0.8, fill = "steelblue", color = "black") +

  scale_x_continuous(
    breaks = x_mapping$x_pos[x_mapping$StateDesc %in% highlight_states],
    labels = x_mapping$StateDesc[x_mapping$StateDesc %in% highlight_states]) +
  
  labs(title = "COPD Rates by City: Highlighting Extremes by State",
       x = "State (ordered by median COPD)",
       y = "COPD Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1)) +
  scale_x_discrete(limits = levels(health$StateDesc), labels = function(x) ifelse(x %in% highlight_states, x, ""))  # Reverses the order on the x-axis

```


Find Healthiest States

```{r}

topTen <- state_means %>%
  arrange(avgMentalIllness) %>%
  head(5)

topTen_dt <- as.data.table(topTen)

healthiest <- data.table(StateDesc = topTen_dt$StateDesc)

names(healthiest)[1] <- "Mental Illness"

topTen <-  state_means %>%
  arrange(avgBingeDrinking) %>%
  head(5)

healthiest$BingeDrinking <- topTen$StateDesc


topTen <-  state_means %>%
  arrange(avgChildhoodAsthma) %>%
  head(5)

healthiest$ChildhoodAsthma <- topTen$StateDesc


topTen <-  state_means %>%
  arrange(avgArthritis) %>%
  head(5)

healthiest$Arthritis <- topTen$StateDesc


topTen <-  state_means %>%
  arrange(avgCancer) %>%
  head(5)

healthiest$Cancer <- topTen$StateDesc


topTen <-  state_means %>%
  arrange(avgKidneyDisease) %>%
  head(5)

healthiest$KidneyDisease <- topTen$StateDesc


topTen <-  state_means %>%
  arrange(avgHighBP) %>%
  head(5)

healthiest$HighBloodPressure <- topTen$StateDesc


topTen <-  state_means %>%
  arrange(avgDiabetes) %>%
  head(5)

healthiest$Diabetes <- topTen$StateDesc


topTen <-  state_means %>%
  arrange(avgCOPD) %>%
  head(5)

healthiest$COPD <- topTen$StateDesc



healthiest$Health <- rep("↓")
healthiest$Health[1] <- "Lowest Rate"
healthiest$Health[5] <- "Fifth Lowest Rate"

healthiest <- healthiest %>%
  select(Health, everything())


```


Make a table for the healthiest places
# ```{R}
# 
# 
# value_counts <- healthiest %>%
#   pivot_longer(everything()) %>%
#   count(value, name = "freq")
# 
# # Step 2: Create a lookup table for frequency
# freq_lookup <- setNames(value_counts$freq, value_counts$value)
# 
# gt_table <- healthiest %>%
#   gt() %>%
#   tab_header(
#     title = "Healthiest States",
#     subtitle = "States with lowest health ascending by Category"
#   ) %>%
#   tab_style(
#     style = cell_text(weight = "bold"),
#     locations = cells_column_labels(columns = everything())
#   ) %>%
#   tab_options(table.width = pct(100),
#   table.font.size = 11,) %>%
#   opt_align_table_header(align = "center") %>%
#   tab_source_note(
#     source_note = md(
#       "**Legend:** Green = appears 5 times &nbsp;&nbsp;&nbsp; Blue = appears 6 times &nbsp;&nbsp;"
#     )
#   )
# 
# for (col_name in colnames(healthiest)) {
#   for (row_index in seq_len(nrow(healthiest))) {
#     cell_value <- healthiest[[col_name]][row_index]
#     freq <- freq_lookup[[cell_value]]
# 
#     if (freq == 5) {
#       gt_table <- gt_table %>%
#         tab_style(
#           style = cell_fill(color = "lightgreen"),
#           locations = cells_body(columns = all_of(col_name), rows = row_index)
#         )
#     }
#     else if (freq == 6) {
#       gt_table <- gt_table %>%
#         tab_style(
#           style = cell_fill(color = "lightblue"),
#           locations = cells_body(columns = all_of(col_name), rows = row_index)
#         )
#     }
#   }
# }
# 
# gt_table
# 
# ```

Make a chart
```{R}

smallest_values <- state_means %>%
  reframe(across(everything(), ~ sort(.x)[1:5])) %>%
  pivot_longer(cols = where(is_double), 
             names_to = "Variable", 
             values_to = "value") %>%
  group_by(Variable) %>%
  mutate(Rank = row_number()) %>%
  ungroup()

ggplot(smallest_values, aes(x = Rank, y = value, color = Variable)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Five Smallest Values per Illness",
    x = "Rank (1 = smallest, 5 = 5th smallest)",
    y = "Value"
  ) +
  theme_minimal() +
  scale_y_continuous(trans='log10') +
  scale_x_continuous(breaks = 1:5)


```












Find Least Healthy States

```{r}

bottomTen <- state_means %>%
  arrange(desc(avgMentalIllness)) %>%
  head(5)

bottomTen_dt <- as.data.table(bottomTen)

leastHealthy <- data.table(StateDesc = bottomTen_dt$StateDesc)

names(leastHealthy)[1] <- "Mental Illness"

bottomTen <-  state_means %>%
  arrange(desc(avgBingeDrinking)) %>%
  head(5)

leastHealthy$BingeDrinking <- bottomTen$StateDesc


bottomTen <-  state_means %>%
  arrange(desc(avgChildhoodAsthma)) %>%
  head(5)

leastHealthy$ChildhoodAsthma <- bottomTen$StateDesc


bottomTen <-  state_means %>%
  arrange(desc(avgArthritis)) %>%
  head(5)

leastHealthy$Arthritis <- bottomTen$StateDesc


bottomTen <-  state_means %>%
  arrange(desc(avgCancer)) %>%
  head(5)

leastHealthy$Cancer <- bottomTen$StateDesc


bottomTen <-  state_means %>%
  arrange(desc(avgKidneyDisease)) %>%
  head(5)

leastHealthy$KidneyDisease <- bottomTen$StateDesc


bottomTen <-  state_means %>%
  arrange(desc(avgHighBP)) %>%
  head(5)

leastHealthy$HighBloodPressure <- bottomTen$StateDesc


bottomTen <-  state_means %>%
  arrange(desc(avgDiabetes)) %>%
  head(5)

leastHealthy$Diabetes <- bottomTen$StateDesc


bottomTen <-  state_means %>%
  arrange(desc(avgCOPD)) %>%
  head(5)

leastHealthy$COPD <- bottomTen$StateDesc


leastHealthy$Health <- rep("↓")
leastHealthy$Health[1] <- "Highest Rate"
leastHealthy$Health[5] <- "Fifth Highest Rate"

leastHealthy <- leastHealthy %>%
  select(Health, everything())


```


# ```{R}
# library(gt)
# 
# value_counts <- leastHealthy %>%
#   pivot_longer(everything()) %>%
#   count(value, name = "freq")
# 
# # Step 2: Create a lookup table for frequency
# freq_lookup <- setNames(value_counts$freq, value_counts$value)
# 
# gt_table <- leastHealthy %>%
#   gt() %>%
#   tab_header(
#     title = "Least Healthy States",
#     subtitle = "States with lowest health ascending by Category"
#   ) %>%
#   tab_options(table.width = pct(100),
#   table.font.size = 11,) %>%
#   opt_align_table_header(align = "center") %>%
#   tab_source_note(
#     source_note = md(
#       "**Legend:** Orange = appears 5 times &nbsp;&nbsp;&nbsp; Yellow = appears 6 times &nbsp;&nbsp;"
#     )
#   )%>%
#   tab_style(
#     style = cell_text(weight = "bold"),
#     locations = cells_column_labels(columns = everything())
#   )
# 
# for (col_name in colnames(leastHealthy)) {
#   for (row_index in seq_len(nrow(leastHealthy))) {
#     cell_value <- leastHealthy[[col_name]][row_index]
#     freq <- freq_lookup[[cell_value]]
# 
#     if (freq == 5) {
#       gt_table <- gt_table %>%
#         tab_style(
#           style = cell_fill(color = "orange"),
#           locations = cells_body(columns = all_of(col_name), rows = row_index)
#         )
#     }
#     else if (freq == 6) {
#       gt_table <- gt_table %>%
#         tab_style(
#           style = list(cell_fill(color = "yellow"),
#                        cell_text(weight = "bold")),
#           locations = cells_body(columns = all_of(col_name), rows = row_index)
#         )
#     }
#   }
# }
# 
# gt_table
# 
# ```




Make a chart
```{R}

biggest_values <- state_means %>%
  reframe(across(everything(), ~ sort(.x, decreasing = TRUE)[1:5])) %>%
  pivot_longer(cols = where(is_double), 
             names_to = "Variable", 
             values_to = "value") %>%
  group_by(Variable) %>%
  mutate(Rank = row_number()) %>%
  ungroup()

ggplot(biggest_values, aes(x = Rank, y = value, color = Variable)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Five Biggest Values per Illness",
    x = "Rank (1 = smallest, 5 = 5th smallest)",
    y = "Value"
  ) +
  theme_minimal() +
  scale_y_continuous(trans='log10') +
  scale_x_continuous(breaks = 1:5)


```






Create a new grouping and perform Kruskal-Wallis analysis
```{R}


group1_states <- c("Alabama", "Georgia", "Louisiana")
group2_states <- c("Ohio", "West Virginia", "Pennsylvania")


state_means$Group <- ifelse(state_means$StateDesc %in% group1_states, "Group1",
                      ifelse(state_means$StateDesc %in% group2_states, "Group2", "Group3"))

group_averages <- state_means %>%
  group_by(Group) %>%
  summarise(across(-StateDesc, mean, na.rm = TRUE))

long_data <- state_means %>%
  pivot_longer(cols = -c(StateDesc, Group), names_to = "Illness", values_to = "Rate")


results <- long_data %>%
  group_by(Illness) %>%
  summarise(p_value = kruskal.test(Rate ~ Group)$p.value)

results <- results[-10, ]


ggplot(results, aes(x = Illness, y = p_value)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  labs(title = "Kruskal-Wallis p-values by Illness",
       x = "Illness",
       y = "p-value") +
  geom_text(aes(label = sprintf("%.4f", p_value)), vjust = -0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

```


```{R}


group1_states <- c("Alabama")


state_means$Group <- ifelse(state_means$StateDesc %in% group1_states, "Group1","Group2")

group_averages <- state_means %>%
  group_by(Group) %>%
  summarise(across(-StateDesc, mean, na.rm = TRUE))

long_data <- state_means %>%
  pivot_longer(cols = -c(StateDesc, Group), names_to = "Illness", values_to = "Rate")


results <- long_data %>%
  group_by(Illness) %>%
  summarise(p_value = kruskal.test(Rate ~ Group)$p.value)

results <- results[-10, ]


ggplot(results, aes(x = Illness, y = p_value)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  labs(title = "Kruskal-Wallis p-values by Illness",
       x = "Illness",
       y = "p-value") +
  geom_text(aes(label = sprintf("%.4f", p_value)), vjust = -0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

```



```{R}


group1_states <- c("Georgia")


state_means$Group <- ifelse(state_means$StateDesc %in% group1_states, "Group1","Group2")

group_averages <- state_means %>%
  group_by(Group) %>%
  summarise(across(-StateDesc, mean, na.rm = TRUE))

long_data <- state_means %>%
  pivot_longer(cols = -c(StateDesc, Group), names_to = "Illness", values_to = "Rate")


results <- long_data %>%
  group_by(Illness) %>%
  summarise(p_value = kruskal.test(Rate ~ Group)$p.value)

results <- results[-10, ]


ggplot(results, aes(x = Illness, y = p_value)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  labs(title = "Kruskal-Wallis p-values by Illness",
       x = "Illness",
       y = "p-value") +
  geom_text(aes(label = sprintf("%.4f", p_value)), vjust = -0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

```

```{R}
group1_states <- c("Louisiana")


state_means$Group <- ifelse(state_means$StateDesc %in% group1_states, "Group1","Group2")

group_averages <- state_means %>%
  group_by(Group) %>%
  summarise(across(-StateDesc, mean, na.rm = TRUE))

long_data <- state_means %>%
  pivot_longer(cols = -c(StateDesc, Group), names_to = "Illness", values_to = "Rate")


results <- long_data %>%
  group_by(Illness) %>%
  summarise(p_value = kruskal.test(Rate ~ Group)$p.value)

results <- results[-10, ]


ggplot(results, aes(x = Illness, y = p_value)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  labs(title = "Kruskal-Wallis p-values by Illness",
       x = "Illness",
       y = "p-value") +
  geom_text(aes(label = sprintf("%.4f", p_value)), vjust = -0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

```
```{r}
highest_illness_table <- state_means %>% 
  select(-Group) %>% 
  pivot_longer(cols = -StateDesc, names_to = "Illness", values_to = "Rate") %>% 
  group_by(StateDesc) %>% 
  slice_max(order_by = Rate, n = 1, with_ties = FALSE)
  
kable(highest_illness_table, caption = "Most Common Illness By State")
```

