---
title: "README"
author: "Naomi Mauss and Grace Wu"
date: "2025-04-16"
output: github_document
format:
  html:
    df-print: gt
    page-layout: full
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This imports the cleaned dataset from the initial data creation step in
Data Creation.Rmd

```{r}
health <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQeaKrReyifqjPVoWB2pYHAo54wLlybjEnMZiPP0i6tdpWb-RLqmwrCvhz8HVOBLCjEo_0qYmrnOkEo/pub?output=csv")


healthwfips <-read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTZ4PRD5X0Zcr1rEWwCTusXmef1IkAFgnpPXeJ1dZ3lVpp7iX-ZsbMFCOpkzS-zSA/pub?output=csv")


```

```{R}

library(readr)
library(dplyr)
library(tidycensus)
library(tidyverse)
library(data.table)
library(gt)
library(xgboost)
library(tigris)
library(ggplot2)
library(sf)
library(stringr)
library(rmapshaper)
library(knitr)

```

```{R}

state_means <- healthwfips %>%
  group_by(StateDesc) %>%
  summarise(
    avgMentalIllness = mean(MHLTH, na.rm = TRUE),
    avgBingeDrinking = mean(BINGE, na.rm = TRUE),
    avgChildhoodAsthma = mean(CASTHMA, na.rm = TRUE),
    avgArthritis = mean(ARTHRITIS, na.rm = TRUE),
    avgCancer = mean(CANCER, na.rm = TRUE),
    avgKidneyDisease = mean(KIDNEY, na.rm = TRUE),
    avgHighBP = mean(BPHIGH, na.rm = TRUE),
    avgDiabetes = mean(DIABETES, na.rm = TRUE),
    avgCOPD = mean(COPD, na.rm = TRUE),
    .groups = "drop"
  )

state_means <- state_means %>%
  mutate(StateDesc = tolower(StateDesc))

state_means <- healthwfips %>% group_by(StateDesc) %>%
  summarise(avgMentalIllness = mean(MHLTH), avgBingeDrinking = mean(BINGE), avgChildhoodAsthma = mean(CASTHMA), avgArthritis = mean(ARTHRITIS), avgCancer = mean(CANCER), avgKidneyDisease = mean(KIDNEY), avgHighBP = mean(BPHIGH), avgDiabetes = mean(DIABETES), avgCOPD = mean(COPD))




state_means$StateDesc[state_means$StateDesc == "South Carolin"] <- "South Carolina"
state_means$StateDesc[state_means$StateDesc == "North Carolin"] <- "North Carolina"

```

Prepare the data to map by state

```{r}
#install.packages("viridis")
#library(ggplot2)
#library(dplyr)
#library(maps)      # for map_data
#library(viridis)   # optional: better color scale

# state_means <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRPtd9hIIamvljlqlmKcpm8xzcoSxkiNT4bLA78cgczFQqrSjI1DRrKnNiKR5Ger9CZHL8AEd0bx7OW/pub?output=csv")

state_means$StateDesc <- str_to_title(state_means$StateDesc)

```

```{R}
# Download the shapefile for all states, including Alaska and Hawaii
us_states_sf <- states(cb = TRUE)

# View the structure of the data (optional)
head(us_states_sf)

# Map state names in your data to their respective state abbreviations
state_abbreviations <- c(
  "Alaska" = "AK", "Hawaii" = "HI", "Alabama" = "AL", "Arizona" = "AZ", 
  "Arkansas" = "AR", "California" = "CA", "Colorado" = "CO", "Connecticut" = "CT",
  "Delaware" = "DE", "Florida" = "FL", "Georgia" = "GA", "Hawaii" = "HI", 
  "Idaho" = "ID", "Illinois" = "IL", "Indiana" = "IN", "Iowa" = "IA", 
  "Kansas" = "KS", "Kentucky" = "KY", "Louisiana" = "LA", "Maine" = "ME", 
  "Maryland" = "MD", "Massachusetts" = "MA", "Michigan" = "MI", "Minnesota" = "MN", 
  "Mississippi" = "MS", "Missouri" = "MO", "Montana" = "MT", "Nebraska" = "NE", 
  "Nevada" = "NV", "New Hampshire" = "NH", "New Jersey" = "NJ", "New Mexico" = "NM", 
  "New York" = "NY", "North Carolina" = "NC", "North Dakota" = "ND", "Ohio" = "OH", 
  "Oklahoma" = "OK", "Oregon" = "OR", "Pennsylvania" = "PA", "Rhode Island" = "RI", 
  "South Carolina" = "SC", "South Dakota" = "SD", "Tennessee" = "TN", "Texas" = "TX", 
  "Utah" = "UT", "Vermont" = "VT", "Virginia" = "VA", "Washington" = "WA", 
  "West Virginia" = "WV", "Wisconsin" = "WI", "Wyoming" = "WY"
)

# my_data <- state_means %>%
#   mutate(state = tolower(StateDesc)) %>%
#   mutate(state_abbr = recode(state, !!!state_abbreviations))

# Merge map data with your data
map_data_combined <- left_join(us_states_sf, state_means, by = c("NAME" = "StateDesc"))

```


The left join the health averages data with geographic data from the sf
package. This will let us map the data later.

City Count By State

``` {R}
city_count_by_state <- health %>%
  group_by(StateDesc) %>%
  summarise(city_count = n_distinct(CityName)) %>% arrange(desc(city_count)) # Use n() if duplicates are allowed

# View result
print(city_count_by_state)
```


``` {R}
ggplot(city_count_by_state, aes(x = reorder(StateDesc, city_count), y = city_count)) +
  geom_col() +
  labs(title = "Number of Cities per State",
       x  = NULL,
       y = "Number of Cities") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.05))) +
  coord_flip() +
  theme_void() +
  theme(axis.text.y = element_text(size = 7),
        axis.text.x = element_text(size = 11),
        axis.title = element_text(size = 12))
```


Now we make the maps and box plots

Mental Illness Map

```{R}

# Plot full map with AK/HI shifted
formatting <- tigris::states() %>%
  dplyr::filter(GEOID < "60") %>%
  tigris::shift_geometry()

formatting <- formatting %>%
  left_join(state_means, by = c("NAME" = "StateDesc"))

# plot(formatting$geometry, formatting$avgArthritis)


ggplot(formatting) +
  geom_sf(aes(fill = avgMentalIllness)) +
  scale_fill_viridis_c(
      option = "plasma",
      direction = -1,
      name = "Avg Mental Illness Rate"
    ) +
    labs(
      title = "Average Mental Illness Rate by State",
      subtitle = "Data from CDC PLACES and ACS",
      caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
    ) +
    theme_void() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12),
      plot.caption = element_text(hjust = 1, size = 8),
      legend.position = "right",
      legend.title = element_text(size = 10, face = "bold"),
      legend.text = element_text(size = 8)
   )


```

This chunk prepares geographic data and visualizes the average mental illness rate by state using a choropleth map.

Mental Illness Box Plot Desc

```{R}
stored_means <- data.frame(health %>%
  group_by(StateDesc) %>%
  summarise(mean_COPD = mean(MHLTH, na.rm = TRUE)) %>%
  arrange(mean_COPD))

highlight_states <- c(
  head(stored_means$StateDesc, 5),
  tail(stored_means$StateDesc, 5)
)

health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"))

health$StateDesc <- reorder(health$StateDesc, health$MHLTH, FUN = median)


health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"),
         StateDesc = factor(StateDesc, levels = stored_means$StateDesc))  # Ensure the correct order

x_mapping <- health %>%
  filter(Highlight == "Highlight") %>%
  arrange(COPD) %>%
  mutate(x_pos = as.numeric(factor(StateDesc)))

# Plot
ggplot() +
  # Plot all "Other" states as points (mean) in the middle
  stat_summary(data = filter(health, Highlight == "Other"),
               aes(x = StateDesc, y = MHLTH),
               fun = mean, geom = "point", shape = 21, fill = "gray60", size = 2) +

  # Plot boxplots for highlighted states (top and bottom 5)
  geom_boxplot(data = filter(health, Highlight == "Highlight"),
               aes(x = StateDesc, y = MHLTH),
               width = 0.8, fill = "steelblue", color = "black") +

  scale_x_continuous(
    breaks = x_mapping$x_pos[x_mapping$StateDesc %in% highlight_states],
    labels = x_mapping$StateDesc[x_mapping$StateDesc %in% highlight_states]) +
  
  labs(title = "Mental Illness Rates by City: Highlighting Extremes by State",
       x = "State (ordered by median mental illness)",
       y = "Mental Illness Rate (%)") +
  theme_minimal() +
 theme(axis.text.x = element_text(angle = 70, hjust = 1, size = 11)) +
  scale_x_discrete(limits = levels(health$StateDesc), labels = function(x) ifelse(x %in% highlight_states, x, ""))

```

This chunk highlights states with the highest and lowest average mental illness rates using box plots, with remaining states show as summary points.

Binge Drinking Map

```{R}
ggplot(formatting) +
  geom_sf(aes(fill = avgBingeDrinking))+
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Binge Drinking Rate"
  ) +
  labs(
    title = "Average Binge Drinking Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )
```

This chunk prepares geographic data and visualizes the average binge drinking rate by state using a choropleth map.

Binge Drinking Box Plot Desc

```{R}
stored_means <- health %>%
  group_by(StateDesc) %>%
  summarise(mean_COPD = mean(BINGE, na.rm = TRUE)) %>%
  arrange(mean_COPD)

highlight_states <- c(
  head(stored_means$StateDesc, 5),
  tail(stored_means$StateDesc, 5)
)

health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"))

health$StateDesc <- reorder(health$StateDesc, health$BINGE, FUN = median)


health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"),
         StateDesc = factor(StateDesc, levels = stored_means$StateDesc))  # Ensure the correct order

x_mapping <- health %>%
  filter(Highlight == "Highlight") %>%
  arrange(COPD) %>%
  mutate(x_pos = as.numeric(factor(StateDesc)))

# Plot
ggplot() +
  # Plot all "Other" states as points (mean) in the middle
  stat_summary(data = filter(health, Highlight == "Other"),
               aes(x = StateDesc, y = BINGE),
               fun = mean, geom = "point", shape = 21, fill = "gray60", size = 2) +

  # Plot boxplots for highlighted states (top and bottom 5)
  geom_boxplot(data = filter(health, Highlight == "Highlight"),
               aes(x = StateDesc, y = BINGE),
               width = 0.8, fill = "steelblue", color = "black") +

  scale_x_continuous(
    breaks = x_mapping$x_pos[x_mapping$StateDesc %in% highlight_states],
    labels = x_mapping$StateDesc[x_mapping$StateDesc %in% highlight_states]) +
  
  labs(title = "Binge Drinking Rates by City: Highlighting Extremes by State",
       x = "State (ordered by median binge drinking)",
       y = "Binge Drinking Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 70, hjust = 1, size = 11)) +
  scale_x_discrete(limits = levels(health$StateDesc), labels = function(x) ifelse(x %in% highlight_states, x, ""))

```

This chunk highlights states with the highest and lowest average binge drinking rates using box plots, with remaining states show as summary points.

Asthma Map

```{R}
ggplot(formatting) +
  geom_sf(aes(fill = avgChildhoodAsthma)) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Childhood Asthma Rate"
  ) +
  labs(
    title = "Average Childhood Asthma Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )
```

This chunk prepares geographic data and visualizes the average asthma rate by state using a choropleth map.

Childhood Asthma Box Plot Desc

```{R}
stored_means <- health %>%
  group_by(StateDesc) %>%
  summarise(mean_COPD = mean(CASTHMA, na.rm = TRUE)) %>%
  arrange(mean_COPD)

highlight_states <- c(
  head(stored_means$StateDesc, 5),
  tail(stored_means$StateDesc, 5)
)

health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"))

health$StateDesc <- reorder(health$StateDesc, health$CASTHMA, FUN = median)


health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"),
         StateDesc = factor(StateDesc, levels = stored_means$StateDesc))  # Ensure the correct order

x_mapping <- health %>%
  filter(Highlight == "Highlight") %>%
  arrange(COPD) %>%
  mutate(x_pos = as.numeric(factor(StateDesc)))

# Plot
ggplot() +
  # Plot all "Other" states as points (mean) in the middle
  stat_summary(data = filter(health, Highlight == "Other"),
               aes(x = StateDesc, y = CASTHMA),
               fun = mean, geom = "point", shape = 21, fill = "gray60", size = 2) +

  # Plot boxplots for highlighted states (top and bottom 5)
  geom_boxplot(data = filter(health, Highlight == "Highlight"),
               aes(x = StateDesc, y = CASTHMA),
               width = 0.8, fill = "steelblue", color = "black") +

  scale_x_continuous(
    breaks = x_mapping$x_pos[x_mapping$StateDesc %in% highlight_states],
    labels = x_mapping$StateDesc[x_mapping$StateDesc %in% highlight_states]) +
  
  labs(title = "Childhood Asthma Rates by City: Highlighting Extremes by State",
       x = "State (ordered by median childhood asthma)",
       y = "Childhood Asthma Rate (%)") +
  theme_minimal() +
 theme(axis.text.x = element_text(angle = 70, hjust = 1, size = 12)) +
  scale_x_discrete(limits = levels(health$StateDesc), labels = function(x) ifelse(x %in% highlight_states, x, ""))

```

This chunk highlights states with the highest and lowest average childhood asthma rates using box plots, with remaining states show as summary points.

Arthritis Map

```{R}
ggplot(formatting) +
  geom_sf(aes(fill = avgArthritis)) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Arthritis Rate"
  ) +
  labs(
    title = "Average Arthritis Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )
```

This chunk prepares geographic data and visualizes the average asthma rate by state using a choropleth map.

Arthritis Box Plot Desc

```{R}
stored_means <- health %>%
  group_by(StateDesc) %>%
  summarise(mean_COPD = mean(ARTHRITIS, na.rm = TRUE)) %>%
  arrange(mean_COPD)

highlight_states <- c(
  head(stored_means$StateDesc, 5),
  tail(stored_means$StateDesc, 5)
)

health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"))

health$StateDesc <- reorder(health$StateDesc, health$ARTHRITIS, FUN = median)


health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"),
         StateDesc = factor(StateDesc, levels = stored_means$StateDesc))  # Ensure the correct order

x_mapping <- health %>%
  filter(Highlight == "Highlight") %>%
  arrange(COPD) %>%
  mutate(x_pos = as.numeric(factor(StateDesc)))

# Plot
ggplot() +
  # Plot all "Other" states as points (mean) in the middle
  stat_summary(data = filter(health, Highlight == "Other"),
               aes(x = StateDesc, y = ARTHRITIS),
               fun = mean, geom = "point", shape = 21, fill = "gray60", size = 2) +

  # Plot boxplots for highlighted states (top and bottom 5)
  geom_boxplot(data = filter(health, Highlight == "Highlight"),
               aes(x = StateDesc, y = ARTHRITIS),
               width = 0.8, fill = "steelblue", color = "black") +

  scale_x_continuous(
    breaks = x_mapping$x_pos[x_mapping$StateDesc %in% highlight_states],
    labels = x_mapping$StateDesc[x_mapping$StateDesc %in% highlight_states]) +
  
  labs(title = "Arthritis Rates by City: Highlighting Extremes by State",
       x = "State (ordered by median arthritis)",
       y = "Childhood Arthritis Rate (%)") +
  theme_minimal() +
 theme(axis.text.x = element_text(angle = 70, hjust = 1, size = 12)) +
  scale_x_discrete(limits = levels(health$StateDesc), labels = function(x) ifelse(x %in% highlight_states, x, ""))

```

This chunk highlights states with the highest and lowest average arthritis rates using box plots, with remaining states show as summary points.

Kidney Disease Map

```{R}
ggplot(formatting) +
  geom_sf(aes(fill = avgKidneyDisease)) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Kidney Disease Rate"
  ) +
  labs(
    title = "Average Kidney Disease Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )
```

This chunk prepares geographic data and visualizes the average kidney disease rate by state using a choropleth map.

Kidney Disease Box Plot Desc

```{R}
stored_means <- health %>%
  group_by(StateDesc) %>%
  summarise(mean_COPD = mean(KIDNEY, na.rm = TRUE)) %>%
  arrange(mean_COPD)

highlight_states <- c(
  head(stored_means$StateDesc, 5),
  tail(stored_means$StateDesc, 5)
)

health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"))

health$StateDesc <- reorder(health$StateDesc, health$KIDNEY, FUN = median)


health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"),
         StateDesc = factor(StateDesc, levels = stored_means$StateDesc))  # Ensure the correct order

x_mapping <- health %>%
  filter(Highlight == "Highlight") %>%
  arrange(COPD) %>%
  mutate(x_pos = as.numeric(factor(StateDesc)))

# Plot
ggplot() +
  # Plot all "Other" states as points (mean) in the middle
  stat_summary(data = filter(health, Highlight == "Other"),
               aes(x = StateDesc, y = KIDNEY),
               fun = mean, geom = "point", shape = 21, fill = "gray60", size = 2) +

  # Plot boxplots for highlighted states (top and bottom 5)
  geom_boxplot(data = filter(health, Highlight == "Highlight"),
               aes(x = StateDesc, y = KIDNEY),
               width = 0.8, fill = "steelblue", color = "black") +

  scale_x_continuous(
    breaks = x_mapping$x_pos[x_mapping$StateDesc %in% highlight_states],
    labels = x_mapping$StateDesc[x_mapping$StateDesc %in% highlight_states]) +
  
  labs(title = "Kidney Disease Rates by City: Highlighting Extremes by State",
       x = "State (ordered by median kidney disease)",
       y = "Kidney Disease Rate (%)") +
  theme_minimal() +
 theme(axis.text.x = element_text(angle = 70, hjust = 1, size = 12)) +
  scale_x_discrete(limits = levels(health$StateDesc), labels = function(x) ifelse(x %in% highlight_states, x, ""))

```

This chunk highlights states with the highest and lowest average kidney disease rates using box plots, with remaining states show as summary points.

High Blood Pressure Map

```{R}
ggplot(formatting) +
  geom_sf(aes(fill = avgHighBP)) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg High Blood Pressure Rate"
  ) +
  labs(
    title = "Average High Blood Pressure Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )
```

This chunk prepares geographic data and visualizes the average high blood pressure rate by state using a choropleth map.

High Blood Pressure Box Plot Desc

```{R}
stored_means <- health %>%
  group_by(StateDesc) %>%
  summarise(mean_COPD = mean(BPHIGH, na.rm = TRUE)) %>%
  arrange(mean_COPD)

highlight_states <- c(
  head(stored_means$StateDesc, 5),
  tail(stored_means$StateDesc, 5)
)

health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"))

health$StateDesc <- reorder(health$StateDesc, health$BPHIGH, FUN = median)


health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"),
         StateDesc = factor(StateDesc, levels = stored_means$StateDesc))  # Ensure the correct order

x_mapping <- health %>%
  filter(Highlight == "Highlight") %>%
  arrange(COPD) %>%
  mutate(x_pos = as.numeric(factor(StateDesc)))

# Plot
ggplot() +
  # Plot all "Other" states as points (mean) in the middle
  stat_summary(data = filter(health, Highlight == "Other"),
               aes(x = StateDesc, y = BPHIGH),
               fun = mean, geom = "point", shape = 21, fill = "gray60", size = 2) +

  # Plot boxplots for highlighted states (top and bottom 5)
  geom_boxplot(data = filter(health, Highlight == "Highlight"),
               aes(x = StateDesc, y = BPHIGH),
               width = 0.8, fill = "steelblue", color = "black") +

  scale_x_continuous(
    breaks = x_mapping$x_pos[x_mapping$StateDesc %in% highlight_states],
    labels = x_mapping$StateDesc[x_mapping$StateDesc %in% highlight_states]) +
  
  labs(title = "High Blood Pressure Rates by City: Highlighting Extremes by State",
       x = "State (ordered by median high blood pressure)",
       y = "High Blood Pressure Rate (%)") +
  theme_minimal() +
 theme(axis.text.x = element_text(angle = 70, hjust = 1, size = 12)) +
  scale_x_discrete(limits = levels(health$StateDesc), labels = function(x) ifelse(x %in% highlight_states, x, ""))

```

This chunk highlights states with the highest and lowest average high blood pressure rates using box plots, with remaining states show as summary points.

Cancer Map

```{R}
ggplot(formatting) +
  geom_sf(aes(fill = avgCancer)) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Cancer Rate"
  ) +
  labs(
    title = "Average Cancer Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8))

```

This chunk prepares geographic data and visualizes the average cancer rate by state using a choropleth map.

Cancer Box Plot Desc

```{R}
stored_means <- health %>%
  group_by(StateDesc) %>%
  summarise(mean_COPD = mean(CANCER, na.rm = TRUE)) %>%
  arrange(mean_COPD)

highlight_states <- c(
  head(stored_means$StateDesc, 5),
  tail(stored_means$StateDesc, 5)
)

health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"))

health$StateDesc <- reorder(health$StateDesc, health$CANCER, FUN = median)


health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"),
         StateDesc = factor(StateDesc, levels = stored_means$StateDesc))  # Ensure the correct order

x_mapping <- health %>%
  filter(Highlight == "Highlight") %>%
  arrange(COPD) %>%
  mutate(x_pos = as.numeric(factor(StateDesc)))

# Plot
ggplot() +
  # Plot all "Other" states as points (mean) in the middle
  stat_summary(data = filter(health, Highlight == "Other"),
               aes(x = StateDesc, y = CANCER),
               fun = mean, geom = "point", shape = 21, fill = "gray60", size = 2) +

  # Plot boxplots for highlighted states (top and bottom 5)
  geom_boxplot(data = filter(health, Highlight == "Highlight"),
               aes(x = StateDesc, y = CANCER),
               width = 0.8, fill = "steelblue", color = "black") +

  scale_x_continuous(
    breaks = x_mapping$x_pos[x_mapping$StateDesc %in% highlight_states],
    labels = x_mapping$StateDesc[x_mapping$StateDesc %in% highlight_states]) +
  
  labs(title = "Cancer Rates by City: Highlighting Extremes by State",
       x = "State (ordered by median cancer)",
       y = "Cancer Rate (%)") +
  theme_minimal() +
 theme(axis.text.x = element_text(angle = 70, hjust = 1, size = 12)) +
  scale_x_discrete(limits = levels(health$StateDesc), labels = function(x) ifelse(x %in% highlight_states, x, ""))

```

This chunk highlights states with the highest and lowest average cancer rates using box plots, with remaining states show as summary points.

Diabetes Map

```{R}

ggplot(formatting) +
  geom_sf(aes(fill = avgDiabetes)) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg Diabetes Rate"
  ) +
  labs(
    title = "Average Diabetes Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8))

```

This chunk prepares geographic data and visualizes the average diabetes rate by state using a choropleth map.

Childhood Box Plot Desc

```{R}
stored_means <- health %>%
  group_by(StateDesc) %>%
  summarise(mean_COPD = mean(DIABETES, na.rm = TRUE)) %>%
  arrange(mean_COPD)

highlight_states <- c(
  head(stored_means$StateDesc, 5),
  tail(stored_means$StateDesc, 5)
)

health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"))

health$StateDesc <- reorder(health$StateDesc, health$DIABETES, FUN = median)


health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"),
         StateDesc = factor(StateDesc, levels = stored_means$StateDesc))  # Ensure the correct order

x_mapping <- health %>%
  filter(Highlight == "Highlight") %>%
  arrange(COPD) %>%
  mutate(x_pos = as.numeric(factor(StateDesc)))

# Plot
ggplot() +
  # Plot all "Other" states as points (mean) in the middle
  stat_summary(data = filter(health, Highlight == "Other"),
               aes(x = StateDesc, y = DIABETES),
               fun = mean, geom = "point", shape = 21, fill = "gray60", size = 2) +

  # Plot boxplots for highlighted states (top and bottom 5)
  geom_boxplot(data = filter(health, Highlight == "Highlight"),
               aes(x = StateDesc, y = DIABETES),
               width = 0.8, fill = "steelblue", color = "black") +

  scale_x_continuous(
    breaks = x_mapping$x_pos[x_mapping$StateDesc %in% highlight_states],
    labels = x_mapping$StateDesc[x_mapping$StateDesc %in% highlight_states]) +
  
  labs(title = "Diabetes Rates by City: Highlighting Extremes by State",
       x = "State (ordered by median diabetes)",
       y = "Diabetes Rate (%)") +
  theme_minimal() +
 theme(axis.text.x = element_text(angle = 70, hjust = 1, size = 12)) +
  scale_x_discrete(limits = levels(health$StateDesc), labels = function(x) ifelse(x %in% highlight_states, x, ""))

```

This chunk highlights states with the highest and lowest average diabetes rates using box plots, with remaining states show as summary points.

COPD Map

```{R}
ggplot(formatting) +
  geom_sf(aes(fill = avgCOPD)) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Avg COPD Rate"
  ) +
  labs(
    title = "Average COPD Rate by State",
    subtitle = "Data from CDC PLACES and ACS",
    caption = "Source: CDC | Visualization by Naomi Mauss and Grace Wu"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 1, size = 8),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8)
  )

```

This chunk prepares geographic data and visualizes the average COPD rate by state using a choropleth map.

COPD Box Plot Desc

```{R}
stored_means <- health %>%
  group_by(StateDesc) %>%
  summarise(mean_COPD = mean(COPD, na.rm = TRUE)) %>%
  arrange(mean_COPD)

highlight_states <- c(
  head(stored_means$StateDesc, 5),
  tail(stored_means$StateDesc, 5)
)



health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"))

health$StateDesc <- reorder(health$StateDesc, health$COPD, FUN = median)


health <- health %>%
  mutate(Highlight = ifelse(StateDesc %in% highlight_states, "Highlight", "Other"),
         StateDesc = factor(StateDesc, levels = stored_means$StateDesc))  # Ensure the correct order

x_mapping <- health %>%
  filter(Highlight == "Highlight") %>%
  arrange(COPD) %>%
  mutate(x_pos = as.numeric(factor(StateDesc)))

# Plot
ggplot() +
  # Plot all "Other" states as points (mean) in the middle
  stat_summary(data = filter(health, Highlight == "Other"),
               aes(x = StateDesc, y = COPD),
               fun = mean, geom = "point", shape = 21, fill = "gray60", size = 2) +

  # Plot boxplots for highlighted states (top and bottom 5)
  geom_boxplot(data = filter(health, Highlight == "Highlight"),
               aes(x = StateDesc, y = COPD),
               width = 0.8, fill = "steelblue", color = "black") +

  scale_x_continuous(
    breaks = x_mapping$x_pos[x_mapping$StateDesc %in% highlight_states],
    labels = x_mapping$StateDesc[x_mapping$StateDesc %in% highlight_states]) +
  
  labs(title = "COPD Rates by City: Highlighting Extremes by State",
       x = "State (ordered by median COPD)",
       y = "COPD Rate (%)") +
  theme_minimal() +
 theme(axis.text.x = element_text(angle = 70, hjust = 1, size = 12)) +
  scale_x_discrete(limits = levels(health$StateDesc), labels = function(x) ifelse(x %in% highlight_states, x, ""))  # Reverses the order on the x-axis

```

This chunk highlights states with the highest and lowest average COPD rates using box plots, with remaining states show as summary points.

Find Healthiest States

```{r}

topTen <- state_means %>%
  arrange(avgMentalIllness) %>%
  head(5)

topTen_dt <- as.data.table(topTen)

healthiest <- data.table(StateDesc = topTen_dt$StateDesc)

names(healthiest)[1] <- "Mental Illness"

topTen <-  state_means %>%
  arrange(avgBingeDrinking) %>%
  head(5)

healthiest$BingeDrinking <- topTen$StateDesc


topTen <-  state_means %>%
  arrange(avgChildhoodAsthma) %>%
  head(5)

healthiest$ChildhoodAsthma <- topTen$StateDesc


topTen <-  state_means %>%
  arrange(avgArthritis) %>%
  head(5)

healthiest$Arthritis <- topTen$StateDesc


topTen <-  state_means %>%
  arrange(avgCancer) %>%
  head(5)

healthiest$Cancer <- topTen$StateDesc


topTen <-  state_means %>%
  arrange(avgKidneyDisease) %>%
  head(5)

healthiest$KidneyDisease <- topTen$StateDesc


topTen <-  state_means %>%
  arrange(avgHighBP) %>%
  head(5)

healthiest$HighBloodPressure <- topTen$StateDesc


topTen <-  state_means %>%
  arrange(avgDiabetes) %>%
  head(5)

healthiest$Diabetes <- topTen$StateDesc


topTen <-  state_means %>%
  arrange(avgCOPD) %>%
  head(5)

healthiest$COPD <- topTen$StateDesc



healthiest$Health <- rep("↓")
healthiest$Health[1] <- "Lowest Rate"
healthiest$Health[5] <- "Fifth Lowest Rate"

healthiest <- healthiest %>%
  select(Health, everything())


```

This chunk identifies the top 5 states with the lowest rates for various health conditions such as mental illness, binge drinking, asthma, arthritis, cancer, and more. It stores them in a consolidated table for reporting.

Visualize the Healthiest States by Condition

```{R}

smallest_values <- state_means %>%
  reframe(across(everything(), ~ sort(.x)[1:5])) %>%
  pivot_longer(cols = where(is_double), 
             names_to = "Variable", 
             values_to = "value") %>%
  group_by(Variable) %>%
  mutate(Rank = row_number()) %>%
  ungroup()

ggplot(smallest_values, aes(x = Rank, y = value, color = Variable)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Five Smallest Values per Illness",
    x = "Rank (1 = smallest, 5 = 5th smallest)",
    y = "Value"
  ) +
  theme_minimal() +
  scale_y_continuous(trans='log10') +
  scale_x_continuous(breaks = 1:5)


```

Find Least Healthy States

```{r}

bottomTen <- state_means %>%
  arrange(desc(avgMentalIllness)) %>%
  head(5)

bottomTen_dt <- as.data.table(bottomTen)

leastHealthy <- data.table(StateDesc = bottomTen_dt$StateDesc)

names(leastHealthy)[1] <- "Mental Illness"

bottomTen <-  state_means %>%
  arrange(desc(avgBingeDrinking)) %>%
  head(5)

leastHealthy$BingeDrinking <- bottomTen$StateDesc


bottomTen <-  state_means %>%
  arrange(desc(avgChildhoodAsthma)) %>%
  head(5)

leastHealthy$ChildhoodAsthma <- bottomTen$StateDesc


bottomTen <-  state_means %>%
  arrange(desc(avgArthritis)) %>%
  head(5)

leastHealthy$Arthritis <- bottomTen$StateDesc


bottomTen <-  state_means %>%
  arrange(desc(avgCancer)) %>%
  head(5)

leastHealthy$Cancer <- bottomTen$StateDesc


bottomTen <-  state_means %>%
  arrange(desc(avgKidneyDisease)) %>%
  head(5)

leastHealthy$KidneyDisease <- bottomTen$StateDesc


bottomTen <-  state_means %>%
  arrange(desc(avgHighBP)) %>%
  head(5)

leastHealthy$HighBloodPressure <- bottomTen$StateDesc


bottomTen <-  state_means %>%
  arrange(desc(avgDiabetes)) %>%
  head(5)

leastHealthy$Diabetes <- bottomTen$StateDesc


bottomTen <-  state_means %>%
  arrange(desc(avgCOPD)) %>%
  head(5)

leastHealthy$COPD <- bottomTen$StateDesc


leastHealthy$Health <- rep("↓")
leastHealthy$Health[1] <- "Highest Rate"
leastHealthy$Health[5] <- "Fifth Highest Rate"

leastHealthy <- leastHealthy %>%
  select(Health, everything())






least_long <- leastHealthy %>%
  select(-1) %>%
  pivot_longer(cols = everything(), values_to = "state") %>%
  count(state) %>%
  mutate(type = "Least Healthy")

most_long <- healthiest %>%
  select(-1) %>%
  pivot_longer(cols = everything(), values_to = "state") %>%
  count(state) %>%
  mutate(type = "Most Healthy")

# Step 2: Combine both
combined_counts <- bind_rows(least_long, most_long)

combined_counts <- combined_counts %>% filter(n > 2)

# Step 3: Plot
ggplot(combined_counts, aes(x = reorder(state, n), y = n, fill = type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Least Healthy" = "orange", "Most Healthy" = "green")) +
  labs(
    title = "Frequency of States in Least/Most Healthy Rankings",
    x = "State",
    y = "Frequency",
    fill = "Health Category"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```

This chunk identifies the top 5 states with the highest rates for various health conditions such as mental illness, binge drinking, asthma, arthritis, cancer, and more. It stores them in a consolidated table for reporting.

Visualize the Least Healthy States by Condition

```{R}

biggest_values <- state_means %>%
  reframe(across(everything(), ~ sort(.x, decreasing = TRUE)[1:5])) %>%
  pivot_longer(cols = where(is_double), 
             names_to = "Variable", 
             values_to = "value") %>%
  group_by(Variable) %>%
  mutate(Rank = row_number()) %>%
  ungroup()

ggplot(biggest_values, aes(x = Rank, y = value, color = Variable)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Five Biggest Values per Illness",
    x = "Rank (1 = smallest, 5 = 5th smallest)",
    y = "Value"
  ) +
  theme_minimal() +
  scale_y_continuous(trans='log10') +
  scale_x_continuous(breaks = 1:5)


```

Create a new grouping and perform Kruskal-Wallis analysis

```{R}


group1_states <- c("Alabama", "Mississippi", "Louisiana")
group2_states <- c("Ohio", "West Virginia", "Pennsylvania")


state_means$Group <- ifelse(state_means$StateDesc %in% group1_states, "Group1",
                      ifelse(state_means$StateDesc %in% group2_states, "Group2", "Group3"))

group_averages <- state_means %>%
  group_by(Group) %>%
  summarise(across(-StateDesc, mean, na.rm = TRUE))

long_data <- state_means %>%
  pivot_longer(cols = -c(StateDesc, Group), names_to = "Illness", values_to = "Rate")


results <- long_data %>%
  group_by(Illness) %>%
  summarise(p_value = kruskal.test(Rate ~ Group)$p.value)

results <- results[-10, ]


ggplot(results, aes(x = Illness, y = p_value)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  labs(title = "Kruskal-Wallis p-values by Illness for Regional Groupings",
       x = "Illness",
       y = "p-value") +
  geom_text(aes(label = sprintf("%.4f", p_value)), vjust = -0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

```

We can see here that many of these values fall well below the red line
at 0.05 that denotes the point at which we decide the p-value is
generally considered statistically significant. Furthermore, arthritis,
COPD, Diabetes, Kidney Disease, and Mental illness all have a p-value of
less than 0.01, which shows that the groupings of states which we chose
are statistically significant in how prominent the various illnesses we
chose are.

```{R}


group1_states <- c("Alabama")


state_means$Group <- ifelse(state_means$StateDesc %in% group1_states, "Group1","Group2")

group_averages <- state_means %>%
  group_by(Group) %>%
  summarise(across(-StateDesc, mean, na.rm = TRUE))

long_data <- state_means %>%
  pivot_longer(cols = -c(StateDesc, Group), names_to = "Illness", values_to = "Rate")


results <- long_data %>%
  group_by(Illness) %>%
  summarise(p_value = kruskal.test(Rate ~ Group)$p.value)

results <- results[-10, ]


ggplot(results, aes(x = Illness, y = p_value)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  labs(title = "Kruskal-Wallis p-values by Illness for Alabama",
       x = "Illness",
       y = "p-value") +
  geom_text(aes(label = sprintf("%.4f", p_value)), vjust = -0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

```

```{R}


group1_states <- c("Mississippi")


state_means$Group <- ifelse(state_means$StateDesc %in% group1_states, "Group1","Group2")

group_averages <- state_means %>%
  group_by(Group) %>%
  summarise(across(-StateDesc, mean, na.rm = TRUE))

long_data <- state_means %>%
  pivot_longer(cols = -c(StateDesc, Group), names_to = "Illness", values_to = "Rate")


results <- long_data %>%
  group_by(Illness) %>%
  summarise(p_value = kruskal.test(Rate ~ Group)$p.value)

results <- results[-10, ]


ggplot(results, aes(x = Illness, y = p_value)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  labs(title = "Kruskal-Wallis p-values by Illness for Mississippi",
       x = "Illness",
       y = "p-value") +
  geom_text(aes(label = sprintf("%.4f", p_value)), vjust = -0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

```

```{R}
group1_states <- c("Louisiana")


state_means$Group <- ifelse(state_means$StateDesc %in% group1_states, "Group1","Group2")

group_averages <- state_means %>%
  group_by(Group) %>%
  summarise(across(-StateDesc, mean, na.rm = TRUE))

long_data <- state_means %>%
  pivot_longer(cols = -c(StateDesc, Group), names_to = "Illness", values_to = "Rate")


results <- long_data %>%
  group_by(Illness) %>%
  summarise(p_value = kruskal.test(Rate ~ Group)$p.value)

results <- results[-10, ]


ggplot(results, aes(x = Illness, y = p_value)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  labs(title = "Kruskal-Wallis p-values by Illness for Louisiana",
       x = "Illness",
       y = "p-value") +
  geom_text(aes(label = sprintf("%.4f", p_value)), vjust = -0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

```

```{R}
group1_states <- c("Ohio")


state_means$Group <- ifelse(state_means$StateDesc %in% group1_states, "Group1","Group2")

group_averages <- state_means %>%
  group_by(Group) %>%
  summarise(across(-StateDesc, mean, na.rm = TRUE))

long_data <- state_means %>%
  pivot_longer(cols = -c(StateDesc, Group), names_to = "Illness", values_to = "Rate")


results <- long_data %>%
  group_by(Illness) %>%
  summarise(p_value = kruskal.test(Rate ~ Group)$p.value)

results <- results[-10, ]


ggplot(results, aes(x = Illness, y = p_value)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  labs(title = "Kruskal-Wallis p-values by Illness for Ohio",
       x = "Illness",
       y = "p-value") +
  geom_text(aes(label = sprintf("%.4f", p_value)), vjust = -0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

```

```{R}
group1_states <- c("West Virginia")


state_means$Group <- ifelse(state_means$StateDesc %in% group1_states, "Group1","Group2")

group_averages <- state_means %>%
  group_by(Group) %>%
  summarise(across(-StateDesc, mean, na.rm = TRUE))

long_data <- state_means %>%
  pivot_longer(cols = -c(StateDesc, Group), names_to = "Illness", values_to = "Rate")


results <- long_data %>%
  group_by(Illness) %>%
  summarise(p_value = kruskal.test(Rate ~ Group)$p.value)

results <- results[-10, ]


ggplot(results, aes(x = Illness, y = p_value)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  labs(title = "Kruskal-Wallis p-values by Illness for West Virginia",
       x = "Illness",
       y = "p-value") +
  geom_text(aes(label = sprintf("%.4f", p_value)), vjust = -0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

```

West Virginia has the lowest p-values across the board, indicating that,
of the states with poor health, it is the most statistically
significant. None of its p-values are below 0.05, but it's still
important to note how much closer it is compared to the other states. Of
course, the Kruskal-Wallis test isn't necessarily intended to be used in
this manner, so it's a pretty rough outline.

```{R}
group1_states <- c("Pennsylvania")


state_means$Group <- ifelse(state_means$StateDesc %in% group1_states, "Group1","Group2")

group_averages <- state_means %>%
  group_by(Group) %>%
  summarise(across(-StateDesc, mean, na.rm = TRUE))

long_data <- state_means %>%
  pivot_longer(cols = -c(StateDesc, Group), names_to = "Illness", values_to = "Rate")


results <- long_data %>%
  group_by(Illness) %>%
  summarise(p_value = kruskal.test(Rate ~ Group)$p.value)

results <- results[-10, ]


ggplot(results, aes(x = Illness, y = p_value)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  labs(title = "Kruskal-Wallis p-values by Illness for Pennsylvania",
       x = "Illness",
       y = "p-value") +
  geom_text(aes(label = sprintf("%.4f", p_value)), vjust = -0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

```

```{r}
highest_illness_table <- state_means %>% 
  select(-Group) %>% 
  pivot_longer(cols = -StateDesc, names_to = "Illness", values_to = "Rate") %>% 
  group_by(StateDesc) %>% 
  slice_max(order_by = Rate, n = 1, with_ties = FALSE)

highest_illness_table <- highest_illness_table %>%
  mutate(Illness = recode(Illness,
                             "avgHighBP" = "High BP  "))
  
kable(highest_illness_table, caption = "Most Common Illness By State")
```

Here, we looked at which illness was the most common for each state. We
can see that across the board, the most common illness is high blood
pressure.

This chunk creates a line chart to compare the five highest values for
each health condition across states. It uses a log scale for clearer
separation of small differences.

Find Correlation between the number of states and

```{R}
vals1 <- unlist(healthiest[ , -1], use.names = FALSE)
vals2 <- unlist(leastHealthy[ , -1], use.names = FALSE)

df1 <- as.data.frame(table(vals1), stringsAsFactors = FALSE)
df2 <- as.data.frame(table(vals2), stringsAsFactors = FALSE)

# Rename columns for clarity
colnames(df1) <- c("StateDesc", "count_df1")
colnames(df2) <- c("StateDesc", "count_df2")


combined <- merge(df1, df2, by = "StateDesc", all = TRUE)


combined[is.na(combined)] <- 0

combined <- merge(combined, city_count_by_state, by = "StateDesc", all = TRUE)

combined$sum_count <- combined$count_df1 + combined$count_df2

ggplot(combined, aes(x = sum_count, y = city_count)) +
  geom_point(size = 3, color = "blue") +
  labs(
    x = "Sum of Extreme Counts (count_df1 + count_df2)",
    y = "Number of Cities",
    title = "Relationship Between Extreme Counts and Number of Cities"
  ) +
  theme_minimal() +
  scale_y_continuous(trans='log10')
```


```{R}
model <- lm(city_count ~ sum_count, data = combined)
summary(model)$r.squared
```

The dotplot I made, as well as the calculated r-squared value of 0.063
shows that there is a low correlation between the number of cities in a
state and the number of times a given state appears in either the top or
bottom 5 healthiest states.



```{r}
numeric_data <- state_means[sapply(state_means, is.numeric)]
cor_matrix <- cor(numeric_data, use = "complete.obs")
cor_melted <- reshape2::melt(cor_matrix)
ggplot(data = cor_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                       midpoint = 0.2, limit = c(-1,1), space = "Lab",
                       name="Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, 
                                   size = 10, hjust = 1)) +
  coord_fixed()
```

There is a strong negative correlation between binge drinking and all
the other variables, as well as a very low correlation between cancer
and all the other variables. There's also a strong correlation between
kidney disease and diabetes, as well as kidney disease and high blood
pressure.

### Conclusion

The most healthy states to live in are Utah, Vermont, North Dakota, and
Minnesota. The least healthy states are West Virginia, Ohio, Delaware,
and Louisiana. West Virginia is significantly less healthy than the rest
of the nation. For further research, we could examine why West Virginia
is so unhealthy, and why some sections of the country are significantly
less healthy than others.

Additionally, we discovered that childhood asthma is very common in
Appalachia. For further research, we could examine if there are any
cultural causes for higher rates of asthma.
